Level-Set method on \texttt{p4est}

### 2015-01-09 paper/tex/applications.tex
Finished my pass

We propose to apply our approach to the study of the phase transition of a liquid melt to a solid crystaline structure. In the case of a single component melt and in the absence of convection, the process is dominated by diffusion and can be modeled as a Stefan problem. We decompose the computational domain $\Omega$ into two subdomains $\Omega^-$ and $\Omega^+$, separated by an interface $\Gamma$. The Stefan problem describes the evolution of the temperature $T$, decomposed into $T_s$ in the solid phase $\Omega^+$ and $T_l$ in the liquid phase $\Omega^-$, as
\begin{align}
\pd{T_l}{t} & = D_l \Delta T_l \quad \mathrm{in} ~~ \Omega^-, \label{eq:stefan_heat_equation_l} \\
\pd{T_s}{t} & = D_s \Delta T_s \quad \mathrm{in} ~~ \Omega^+. \label{eq:stefan_heat_equation_s}
\end{align}
The diffusion constants $D_l$ and $D_s$ can be discontinuous across the interface. We prescribe homogeneous Neumann boundary conditions on the edge of the computational domain, $\nabla T \cdot \mathbf{n}\vert_{\partial \Omega}=0$. At the interface between the solid and the liquid phases, the temperature is given by the Gibbs-Tompson boundary condition \cite{Alexiades;Solomon;Wilson:88:The-formation-of-a-s, Alexiades;Solomon:93:Mathematical-Modelin}
\begin{equation} \label{eq:stefan_gibbs_tompson}
T_s = T_l = T_{\Gamma} = -\epsilon_c \kappa - \epsilon_v (\mathbf{V} \cdot \mathbf{n}),
\end{equation}
where $\kappa$ is the local interface curvature, $V$ is the velocity of the interface, $\mathbf{n}$ is the outward normal to the solidification front and $\epsilon_c$ and $\epsilon_v$ are the curvature and kinetic undercooling coefficients. The interface velocity $V$ is defined from the jump in the heat flux at the interface,
\begin{equation} \label{eq:stefan_velocity}
(\mathbf{V} \cdot \mathbf{n}) = - \left[ D_l \pd{T_l}{\mathbf{n}} - D_s \pd{T_s}{\mathbf{n}} \right].
\end{equation}
We choose to use an adaptive time step with a $\text{CFL} = 5$, i.e.
\begin{equation} \label{eq:stefan_dt}
\Delta t = 5 ~ \Delta x_{min} ~ \min(1,1/\max \lVert \mathbf{V} \rVert),
\end{equation}
where $\Delta x_{min}$ is the size of the smallest cell of the forest. The general procedure to solve the Stefan problem is presented in algorithm \ref{alg:stefan}.

\begin{algorithm}[htbp]
\caption{$\texttt{General procedure for solving the Stefan problem}$}
\begin{algorithmic}[1]
\State $\texttt{Initialize the forest and }\phi\texttt{ given the initial geometry}.$
\State $\texttt{Initialize }T_s\texttt{ in }\Omega^+\texttt{ and }T_l\texttt{ in }\Omega^-.$
\State $\texttt{Reinitialize }\phi\texttt{ and compute the local interface curvature }\kappa.$
\State $\texttt{Compute }T^{n+1}_l\texttt{ and }T^{n+1}_s\texttt{ by solving the heat equations \eqref{eq:stefan_heat_equation_l} and \eqref{eq:stefan_heat_equation_s}}.$
\State $\texttt{Extrapolate }T^{n+1}_s\texttt{ from }\Omega^+\texttt{ to }\Omega^-\texttt{ and }T^{n+1}_l\texttt{ from }\Omega^-\texttt{ to }\Omega^+.$
\State $\texttt{Compute the velocity field }\mathbf{V}\texttt{ according to \eqref{eq:stefan_velocity}}.$
\State $\texttt{Compute the time step dt following \eqref{eq:stefan_dt}.}$
\State $\texttt{Evolve the interface and contruct the new forest using the Semi-Lagrangian procedure}.$
\State $\texttt{Interpolate }T^{n+1}_s\texttt{ and }T^{n+1}_l\texttt{ from the old forest to the new forest}.$
\State $\texttt{Go to 3 with n=n+1}.$
\end{algorithmic}
\label{alg:stefan}
\end{algorithm}

@@

\begin{figure}
\centering
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_scaling_small.pdf}}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_scaling_large.pdf}}
\caption{Scalability of the Stefan problem for small (left) and large (right) octrees with roughly 7M and 105M grid points, respectively. As expected from the scalability analysis of the individual components, we observe excellent results, illustrating the potential of our solver.}
\label{fig:stefan_scaling}
\end{figure}

\subsection{Numerical experiments}

We now present the results from a large simulation of the Stefan problem. We choose to work with a $20\times20\times20$ macro-mesh and with level-10 octrees. We choose to define the Gibbs-Tompson anisotropy undercooling coefficients \eqref{eq:stefan_gibbs_tompson} as
\begin{align*}
\epsilon_c & = \left[ \epsilon_1 \left( 1+\alpha_1 \cos(3\theta_1) \right) + \epsilon_2 \left( 1+\alpha_2 \cos(3\theta_2) \right) \right] \kappa,\\
\epsilon_v & = 0,
\end{align*}
with $\theta_1$ the angle between the normal to the interface $\mathbf{n}$ and the x-axis in the $(x,y)$ plane, $\theta_2$ the angle between $\mathbf{n}$ and the x-axis in the $(x,z)$ plane, and
\begin{align*}
\epsilon_1 & = 2~(\sin(x)+\cos(y)+2)\cdot10^{-6}, & \epsilon_2 & = 2~(\sin(x)+\cos(z)+2)\cdot10^{-6}, \\
\alpha_1 & = \frac{1}{4}(\cos(x)+\sin(y)+2), & \alpha_2 & = \frac{1}{4}(\cos(x)+\sin(z)+2).
\end{align*}
Note that these expressions are arbitrary and lead to very varied crystal shapes. The computation is initialized with twenty spherical seeds of radius $0.0015$ placed randomly in the domain. We choose $D_s=D_l=1$ and set the initial temperatures $T^0_l=-0.25$ and $T^0_s=0$.

We let the simulation run with $256$ mpi processes for $6$ hours and $30$ minutes, resulting in $396$ time iterations. Visualizations of the final iteration are presented in figures \ref{fig:stefan_grid} and \ref{fig:stefan_crystals}. The final iteration of the simulation consisted of $167$M grid points when a uniform grid with the equivalent finest resolution would lead to $8.59\cdot10^{12}$ grid points, i.e. over eight trillion grid points ! Our simulation used only $0.001948\%$ of the number of grid points needed for the same simulation on a uniform grid. This application demonstrates the potential of our approach.

>>>>

We propose to apply \mohammad{what about: In this section we apply ...} our approach to the study of the phase transition of a liquid melt to a solid crystaline structure. In the case of a single component melt, and in the absence of convection, the process is dominated by diffusion and can be modeled as a Stefan problem. We decompose the computational domain $\Omega$ into \mohammad{How about: Consider a computational domain $\Omega$ decomposed into ...} two subdomains $\Omega^-$ and $\Omega^+$, separated by an interface $\Gamma$. The Stefan problem describes the evolution of the temperature $T$, decomposed into $T_s$ in the solid phase $\Omega^+$ and $T_l$ in the liquid phase $\Omega^-$, as
\begin{align}
\pd{T_l}{t} & = D_l \Delta T_l \quad \mathrm{in} ~~ \Omega^-, \label{eq:stefan_heat_equation_l} \\
\pd{T_s}{t} & = D_s \Delta T_s \quad \mathrm{in} ~~ \Omega^+. \label{eq:stefan_heat_equation_s}
\end{align}
The diffusion constants $D_l$ and $D_s$ can be discontinuous across the interface. We prescribe homogeneous Neumann boundary conditions on the edge of the computational domain, $\nabla T \cdot \mathbf{n}\vert_{\partial \Omega}=0$. At the interface between the solid and the liquid phases, the temperature is given by the Gibbs-Tompson boundary condition \cite{Alexiades;Solomon;Wilson:88:The-formation-of-a-s, Alexiades;Solomon:93:Mathematical-Modelin}
\begin{equation} \label{eq:stefan_gibbs_tompson}
T_s = T_l = T_{\Gamma} = -\epsilon_c \kappa - \epsilon_v (\mathbf{V} \cdot \mathbf{n}),
\end{equation}
where $\kappa$ is the local interface curvature, $V$ is the velocity of the interface, $\mathbf{n}$ is the outward normal to the solidification front and $\epsilon_c$ and $\epsilon_v$ are the curvature and kinetic undercooling coefficients. The interface velocity $V$ is defined from the jump in the heat flux at the interface,
\begin{equation} \label{eq:stefan_velocity}
(\mathbf{V} \cdot \mathbf{n}) = - \left[ D_l \pd{T_l}{\mathbf{n}} - D_s \pd{T_s}{\mathbf{n}} \right].
\end{equation}
We choose to use an adaptive time step with a $\text{CFL} = 5$, i.e.
\begin{equation} \label{eq:stefan_dt}
\Delta t = 5 ~ \Delta x_{min} ~ \min(1,1/\max \lVert \mathbf{V} \rVert),
\end{equation}
where $\Delta x_{min}$ is the size of the smallest cell of the forest. The general procedure to solve the Stefan problem is presented in algorithm \ref{alg:stefan}.

\begin{algorithm}[htbp]
\caption{$\texttt{General procedure for solving the Stefan problem}$}
\mohammad{Is there a reason all of the algorithm is in type-writer font? Usually its reserved for code segments, etc.}
\begin{algorithmic}[1]
\State $\texttt{Initialize the forest and }\phi\texttt{ given the initial geometry}.$
\State $\texttt{Initialize }T_s\texttt{ in }\Omega^+\texttt{ and }T_l\texttt{ in }\Omega^-.$
\State $\texttt{Reinitialize }\phi\texttt{ and compute the local interface curvature }\kappa.$
\State $\texttt{Compute }T^{n+1}_l\texttt{ and }T^{n+1}_s\texttt{ by solving the heat equations \eqref{eq:stefan_heat_equation_l} and \eqref{eq:stefan_heat_equation_s}}.$
\State $\texttt{Extrapolate }T^{n+1}_s\texttt{ from }\Omega^+\texttt{ to }\Omega^-\texttt{ and }T^{n+1}_l\texttt{ from }\Omega^-\texttt{ to }\Omega^+.$
\State $\texttt{Compute the velocity field }\mathbf{V}\texttt{ according to \eqref{eq:stefan_velocity}}.$
\State $\texttt{Compute the time step dt following \eqref{eq:stefan_dt}.}$
\State $\texttt{Evolve the interface and contruct the new forest using the Semi-Lagrangian procedure}.$
\State $\texttt{Interpolate }T^{n+1}_s\texttt{ and }T^{n+1}_l\texttt{ from the old forest to the new forest}.$
\State $\texttt{Go to 3 with n=n+1}.$
\end{algorithmic}
\label{alg:stefan}
\end{algorithm}

@@

\begin{figure}
\centering
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_scaling_small.pdf}}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_scaling_large.pdf}}
\caption{Scalability of the Stefan problem for small (left) and large (right) octrees with roughly 7M and 105M grid points, respectively. As expected from the scalability analysis of the individual components, we observe excellent results, illustrating the potential \mohammad{applicability?} of our solver \mohammad{algorithms?}.}
\label{fig:stefan_scaling}
\end{figure}

\subsection{Numerical experiments}
\mohammad{too many transitive statements and keyword ``choose''. Can we rephrase this?}
We now present the results from a large simulation of the Stefan problem. We choose to work with a $20\times20\times20$ macro-mesh and with level-10 octrees. We choose to define the Gibbs-Tompson anisotropy undercooling coefficients \eqref{eq:stefan_gibbs_tompson} as
\begin{align*}
\epsilon_c & = \left[ \epsilon_1 \left( 1+\alpha_1 \cos(3\theta_1) \right) + \epsilon_2 \left( 1+\alpha_2 \cos(3\theta_2) \right) \right] \kappa,\\
\epsilon_v & = 0,
\end{align*}
with $\theta_1$ the angle between the normal to the interface $\mathbf{n}$ and the x-axis in the $(x,y)$ plane, $\theta_2$ the angle between $\mathbf{n}$ and the x-axis in the $(x,z)$ plane, and
\begin{align*}
\epsilon_1 & = 2~(\sin(x)+\cos(y)+2)\cdot10^{-6}, & \epsilon_2 & = 2~(\sin(x)+\cos(z)+2)\cdot10^{-6}, \\
\alpha_1 & = \frac{1}{4}(\cos(x)+\sin(y)+2), & \alpha_2 & = \frac{1}{4}(\cos(x)+\sin(z)+2).
\end{align*}
Note that these expressions are arbitrary and lead to very varied crystal shapes \mohammad{How about: ... lead to a variety of shapes?}. The computation is initialized with twenty spherical seeds of radius $0.0015$ placed randomly in the domain. We choose $D_s=D_l=1$ and set the initial temperatures $T^0_l=-0.25$ and $T^0_s=0$.

We let the simulation run with \mohammad{How about: The simulation is ran on ...} $256$ mpi processes for $6$ hours and $30$ minutes, resulting in $396$ time iterations. Visualizations of the final iteration are presented in figures \ref{fig:stefan_grid} and \ref{fig:stefan_crystals}. The final iteration of the simulation consisted of $167$M grid points when a uniform grid with the equivalent finest resolution would lead to $8.59\cdot10^{12}$ grid points, i.e. over eight trillion grid points ! Our simulation used only $0.001948\%$ \mohammad{I think if we are going to report percentage, we should trim it to 1st digit} of the number of grid points needed for the same simulation on a uniform grid. This application demonstrates the potential of our approach.

### 2015-01-09 paper/tex/conclusion.tex
Finished my pass

%!TEX root = draft.tex
\section{Conclusions}
In this article we have presented parallel algorithms for the advection and reinitialization of the level-set functions on adaptive Quadtree and Octree grids using domain decomposition approach. These algorithms are implemented using a combination of \texttt{MPI} standard and the open-source \texttt{p4est} library. An important feature of the Semi-Lagrangian method is its unconditional stability property which must be preserved in parallel, requiring a parallel interpolation scheme. However, due to the non-uniform distribution of departure points, scalable implementation of such algorithm is not trivial on adaptive Quadtree and Octree grids. An asynchronous interpolation algorithm is presented using non-blocking point-to-point communications which proves good scalability.

The scalability of the Semi-Lagrangian algorithm, however, depends on the CFL number. Great scalability is observed for intermediate CFL numbers while, e.g. $\text{CFL}~10$. At higher CFL numbers, however, the departure points are potentially further dispersed across processors which limits the scalability. This is due to the fact the domain decomposition technique used here is based on the Z-ordering of quadrants and does not take the velocity field information into account. A possible remedy for this problem could be assigning weights to quadrants based on some estimate of what the grid should look like after one step of the advection algorithm, e.g. by using a forward-in-time integration of grid points. Such an estimate could also reduce the number of Semi-Lagrangian iterations. These ideas are postponed for further investigations. We have also presented a simple parallelization technique for the reinitialization algorithm based on the pseudo-time transient formulation. Both the Semi-Lagrangian and the reinitialization algorithms show great scalability up to 4096 processors.

Finally, an application of these algorithms is presented in modeling the solidification process by solving the Stefan problem. This application clearly illustrates the applicability of our algorithms to complex multi-scale problems that cannot be treated using the normal domain decomposition techniques on uniform grids. We believe that our findings would be of interest to other researchers using the level-set framework for other complex and multi-scale problems.

\section*{Acknowledgment} 
The funding for this project was provided by \comment{the funding agency}. We would also like to thank the technical staff at the Texas Advanced Computing Center (TACC) and the developer community of PETSc library for their valuable suggestions.

>>>>

%!TEX root = draft.tex
\section{Conclusions}
In this article we have presented parallel algorithms for the advection and reinitialization of the level-set functions on adaptive Quadtree and Octree grids using domain decomposition approach. These algorithms are implemented using a combination of \texttt{MPI} and the open-source \texttt{p4est} library. An important feature of the semi-Lagrangian method is its unconditional stability property which must be preserved in parallel, requiring a parallel interpolation scheme. However, due to the non-uniform distribution of departure points, scalable implementation of such algorithm is not trivial on adaptive Quadtree and Octree grids. An asynchronous interpolation algorithm is presented using non-blocking point-to-point communications which proves good scalability.

The scalability of the semi-Lagrangian algorithm, however, depends on the CFL number. Great scalability is observed for intermediate CFL numbers while, e.g. $\text{CFL}\sim10$. At higher CFL numbers, however, the departure points are potentially further dispersed across processors which limits the scalability. This is because the domain decomposition technique used here is based on the Z-ordering of cells and does not take the velocity field information into account. A possible remedy for this problem could be assigning weights to cells based on some estimate of what the grid should look like after one step of the advection algorithm, e.g. by using a forward-in-time integration of grid points. Such an estimate could also reduce the number of semi-Lagrangian iterations. These ideas are postponed for further investigations. We have also presented a simple parallelization technique for the reinitialization algorithm based on the pseudo-time transient formulation. Both the semi-Lagrangian and the reinitialization algorithms show good scalability up to 4096 processors.

Finally, an application of these algorithms is presented in modeling the solidification process by solving a Stefan problem. This application clearly illustrates the applicability of our algorithms to complex multi-scale problems that cannot be treated using the normal domain decomposition techniques on uniform grids. We believe that our findings would be of interest to other researchers who use the level-set framework for similar complex and multi-scale problems.

\section*{Acknowledgment} 
The funding for this project was provided by \mohammad{the funding agency}. We would also like to thank the technical staff at the Texas Advanced Computing Center (TACC) and the developer community of PETSc library for their valuable suggestions during this project.
